Constructing Biological Regulatory Networks from Process Hitting models

### 2013-07-05 main.tex
new title
typo in title
new section for ASP
several TODOs
fix old inference (remove gamma) and add new inference (new section)
finish new inference
avancées sur l'intro et l'abstract

% Commandes À FAIRE
%\usepackage{color} % Couleurs du texte
\newcommand{\rewrite}[1]{\textcolor{blue}{\textbf{[REWRITE: #1]}}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
\newcommand{\new}[1]{\textcolor{darkgreen}{\textbf{[$^\bigstar$New:} #1\textbf{$^\bigstar$]}}}
%\newcommand{\new}[1]{#1}

@@

\title{\todo{New title} Concretizing the Process Hitting into Biological Regulatory Networks}

@@

\address[irccyn]{LUNAM Universit\'e, \'Ecole Centrale de Nantes, IRCCyN UMR CNRS 6597\\
(Institut de Recherche en Communications et Cybern\'etique de Nantes)\\
1 rue de la No\"e - B.P. 92101 - 44321 Nantes Cedex 3, France.}
\address[nii]{National Institute of Informatics,\\
2-1-2, Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan.}
\address[eth]{ETH Zürich, Institut für Automatik,\\
ETL I 28, Physikstrasse 3, 8092 Zurich, Switzerland.}

@@

Given a PH model of a BRN, we first tackle the inference of the underlying Interaction Graph
between components.
Then the inference of corresponding Thomas' models is provided using Answer Set Programming,
which allows notably an efficient enumeration of (possibly numerous) compatible parametrizations.

In addition to giving a formal link between different approaches for qualitative BRNs modeling, 
this work emphasizes the ability of PH to deal with large BRNs with incomplete knowledge on
cooperations, where Thomas' approach fails because of the combinatorics of parameters.

\todo{Emphasis on the new contribution since CMSB.}
\end{abstract}

@@

\input{parts/infer-ig}
\input{parts/infer-param}
\input{parts/example}
\input{parts/discussion}

>>>>

% Commandes À FAIRE
%\usepackage{color} % Couleurs du texte
\newcommand{\warnul}[1]{\textcolor{red}{\underline{#1}}}
\newcommand{\rewriteng}[1]{\textcolor{blue}{#1}}
\newcommand{\rewriteil}[1]{\rewriteng{\textbf{[#1]}}}
\newcommand{\rewrite}[1]{\rewriteil{REWRITE: #1}}
\newcommand{\towrite}[1]{\rewriteil{TOWRITE: #1}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
\newcommand{\new}[1]{\textcolor{darkgreen}{\textbf{[$^\bigstar$New:} #1\textbf{$^\bigstar$]}}}
\def\updated#1{{\color{magenta}#1}}
%\newcommand{\new}[1]{#1}

@@

\title{Constructing Biological Regulatory Networks from Process Hitting models}
%\\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

@@

\address[irccyn]{LUNAM Universit\'e, \'Ecole Centrale de Nantes, IRCCyN UMR CNRS 6597\\
(Institut de Recherche en Communications et Cybern\'etique de Nantes)\\
1 rue de la No\"e - B.P. 92101 - 44321 Nantes Cedex 3, France.}
\address[nii]{National Institute of Informatics,\\
2-1-2, Hitotsubashi, Chiyoda-ku, Tokyo 101-8430, Japan.}
\address[eth]{BISON group, Automatic Control Laboratory, ETH Zürich\\
Physikstrasse 3, 8092 Zurich, Switzerland.}

@@

This paper establishes formal links between these two formalisms, permitting to appreciate their
complementary features.
First, the inference of the Interaction Graph (IG) from a PH models allows a compact representation of
the regulations that are effective for the dynamics.
From this IG, well-known results can be applied to derive global dynamical properties for the
modelled BRN.
Second, the inference of all Thomas models that are compatible with a given PH is tackled.
As the PH allows partial specification of cooperation between components, this inference
emphasizes the ability of PH to deal with large BRNs with incomplete knowledge on
cooperations, where Thomas' approach fails because of the combinatorics of parameters.

The inference of corresponding Thomas' models is provided using Answer Set Programming,
which allows notably an efficient enumeration of (possibly numerous) compatible parametrizations.
\todo{application}
\end{abstract}

@@

\input{parts/infer-ig}
\input{parts/infer-param}
\input{parts/asp}
\input{parts/example}
\input{parts/discussion}

### 2013-07-05 parts/infer-ig.tex
less spurious auto-edges in IG inference
avancées sur l'intro et l'abstract

\section{Interaction Graph Inference}\label{sec:infer-IG}

In order to infer a complete BRN, one has to find the Interaction Graph (IG) first, as some
constraints on the parametrization rely on it.
Inferring the IG is an abstraction step which consists in determining the global influence of
components on each of its successors.

@@

    &\qquad s = \f{sign}(a_k - a_j)

@@

where $\f{sign}(x) = \begin{cases}\frac{|x|}{x} & \text{ if $x \neq 0$} \\ 0 & \text{ if $x = 0$}\end{cases}$.

@@

@@

We point out that one of the benefits of this inference is the fact that the conditions used to infer each influence
in \pref{pps:inference-edges}
allow to apply some results about the links between the structure of an IG and the dynamics of the model.
These results consist mainly in Thomas' conjectures, that focus on the presence of positive and negative circuits in the IG, and state that:
\begin{itemize}
  \item the presence of a positive circuit in the IG is a necessary condition for several fixed points,
  \item the presence of a negative circuit in the IG is a necessary condition for sustained oscillations.
\end{itemize}
The sign of a circuit is defined by the number of negative edges it contains;
given the nature of unsigned edges, the presence of such an edge in a circuit would count for the presence of both a positive and negative circuit.
These conjectures have been demonstrated in~\cite{Richard2010378} \towrite{Other references! (multi-stationarity \& boolean models)}.

>>>>

% vim:spell spelllang=en:
\section{Interaction Graph Inference from Process Hitting}\label{sec:infer-IG}

\updated{%
The Interaction Graph (IG) is an abstract representation of the direct qualitative influences,
positives and/or negatives, between the components of the system.
As discussed in \pref{sec:intro}, the IG allows to efficiently characterise global dynamical
properties for the concrete system, such as the capability for multi-stationarity or oscillation.}

\updated{%
In a typical biological network modelling process, a prior IG is generally the starting point for
the formal system specification.
However, it is very common that the prior IG actually refers interaction that reveal to be 
non-effective with respect to the dynamics.
Hence, deriving the IG directly from the dynamical models lead to more concise IGs, enhancing the
conclusiveness of static analyses based upon this abstract representation.}

\updated{%
In this section, we formally derive the IG corresponding to a given PH that is well-formed for BRN
modelling.}

%In order to infer a complete BRN, one has to find the Interaction Graph (IG) first, as some
%constraints on the parametrization rely on it.
%Inferring the IG is an abstraction step which consists in determining the global influence of
%components on each of its successors.

@@

    &\qquad s = \f{sign}(k - a)

@@

where $\f{sign}(x) = \begin{cases}|x|/x & \text{ if $x \neq 0$} \\ 0 & \text{ if $x = 0$}\end{cases}$.

@@

\todo{remark for reducing the number of self-regulation}

@@

### 2013-07-05 parts/intro.tex
avancées sur l'intro et l'abstract

\section{Introduction}\label{sec:intro}
As regulatory phenomena play a crucial role in biological systems, they need to be studied accurately.
Biological Regulatory Networks (BRNs) consist in sets of either positive or negative mutual effects between the components.
With the purpose of analyzing these systems, they are often modeled as graphs which make it possible to determine the possible evolutions of all the interacting components of the system.
Indeed, besides continuous models of physicists, often designed through systems of ordinary
differential equations, a discrete modeling approach was initiated by René Thomas in 1973
\cite{Thomas73}.

@@

Thomas' modeling has motivated numerous works around the link between the Interaction Graph (IG)
(summarizing the global influences between components) and the possible dynamics (e.g.,
\cite{RiCo07,RRT08}),
model reduction (e.g., \cite{Naldi09}), formal checking of dynamics (e.g., \cite{Richard06,Naldi07}), 
and the incorporation of time (e.g., \cite{Siebert06,Ahmad08}) and probability
(e.g., \cite{Twardziok10-CMSB}) dimensions, to name but a few.
While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, the main drawback of this framework is the difficulty to specify Thomas'
parameters, especially for large networks.

In order to address the formal checking of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{``Process Hitting''} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, where the causality between actions is restricted.
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of BRNs with hundreds of components \cite{PMR12-MSCS}.

PH is suitable to model BRNs with different levels of abstraction in the specification of
cooperations (associated influences) between components.
This allows to model BRNs with a partial knowledge on precise evolution functions for components
by capturing the largest (the most general) dynamics.

The objectives of the work presented in this paper are the following.
Firstly, we show that starting from one PH model, it is possible to find back the underlying IG.
We perform an exhaustive search for the possible interactions on one component from all the
others, consistently with the knowledge of the dynamics that these interactions lead to and that are
expressed in PH.
The second phase of our work concerns the Thomas' parameters inference.
It consists in determining the nesting set (possibly too large) of the parameters which necessarily
lead to the satisfaction of the known cooperating constraints.
The resulting BRN dynamics is ensured to respect the PH dynamics, \ie no spurious transitions are
made possible by the inference.
Answer Set Programming (ASP) \cite{Baral03} turns out to be effective for these enumerative searches.

The outcome of this work is twofold.
The first benefit is that such an approach makes it possible to refine the construction of
BRNs with a partial and progressively brought knowledge in PH, while being able to export such
models in the Thomas' framework.
This work thus strengthens the formal link between both modelings.
The second feature of our method is that it can be applied on very large BRNs.

Finally, it must be noticed that we are not interested in this paper in the derivation of one
PH from a BRN (which was previously described in \cite{PMR10-TCSB}) but, on the contrary, to finding out
a set of BRNs from one PH.

Our work is related to the approach of \cite{Khalis09} which relies on temporal logic, and \cite{20646302,DBLP:conf/ipcat/CorblinFTCT12} which also uses constraint programming. Both aim at determining a class of models which are consistent with available partial data on the regulatory structure and dynamical properties.
%These classes are built in order to infer properties common to some studied models.
%In our approach, we intend to focus on the Thomas' parameters inference
Our method is based on a model rather than on constraints, which allows to define some properties on the system structure (such as cooperations).
Furthermore, we claim that we are able to deal with larger biological networks.

\todo{Emphasis on the new contribution since CMSB.}

\paragraph{Outline}
\pref{sec:frameworks} recalls the PH and Thomas frameworks;
\pref{sec:infer-IG} defines the IG inference from PH;
\pref{sec:infer-K} details the enumeration of Thomas parametrizations compatible with a PH
and discuss its implementation in ASP.
\pref{sec:examples} illustrates the applicability of our method on simple examples
and large biological models.

>>>>

% vim:spell spelllang=en:
\section{Introduction}\label{sec:intro}

\rewrite{Include Olivier's additions}

As regulatory phenomena play a crucial role in biological systems, they need to be studied accurately.
Biological Regulatory Networks (BRNs) consist in sets of either positive or negative mutual effects between the components.
With the purpose of analyzing these systems, they are often modeled as graphs which make it possible to determine the possible evolutions of all the interacting components of the system.
Indeed, besides continuous models of physicists, often designed through systems of ordinary
differential equations, a discrete modeling approach was initiated by René Thomas in 1973
\cite{Thomas73}.

@@

Thomas' modeling has motivated numerous works around the link between the Interaction Graph (IG)
(summarizing the global influences between components) and the possible dynamics (e.g.,
\cite{RiCo07,RRT08}),
model reduction (e.g., \cite{Naldi09}), formal checking of dynamics (e.g., \cite{Richard06,Naldi07}), 
and the incorporation of time (e.g., \cite{Siebert06,Ahmad08}) and probability
(e.g., \cite{Twardziok10-CMSB}) dimensions, to name but a few.

While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, \updated{an other major} drawback of this framework is the difficulty to
specify Thomas' parameters, especially for large networks.

In order to address the formal analysis of dynamical properties within very large BRNs, we recently
introduced in \cite{PMR10-TCSB} a new formalism, named the \emph{Process Hitting} (PH), to model
concurrent systems having components with a few qualitative levels.
A PH model describes, in an atomic manner, the possible evolutions of a process (representing one
component at one level) triggered by the hit of at most one other process in the system.
This framework can be seen as a special class of formalisms like Petri Nets or Communicating Finite
State Machines, \updated{where the arity and the effect of synchronisations between components is
restricted.}
Thanks to the particular structure of interactions within a PH, very efficient static analysis
methods have been developed to over- and under-approximate reachability properties making tractable
the formal analysis of the qualitative dynamics of BRNs with hundreds or thousands of components,
\updated{which was impossible with other state-of-the-art approaches \cite{PMR12-MSCS,PAK13-CAV,FPMR13-CS2Bio}}.

The PH is suitable to model BRNs with different levels of abstraction in the specification of
cooperations (associated influences) between components.
This allows to model BRNs with a partial knowledge on precise evolution functions for components
by capturing the largest (the most general) dynamics.

\medskip

\updated{%
The aim of this paper is to formally estiblish links between several layers suited for specifying the
qualitative dynamics of BRNs, namely:
the Interaction Graph (IG), referencing the sign of direct regulations between components;
the Process Hitting (PH), which permits to specify the qualitatif dynamics of the system with
various degrees of knowledge for joint regulations;
and Thomas' parameters, which completely specify the functions governing the evolution of
components.}

\updated{%
Firstly, we derive rules that, given a PH model, infer the IG corresponding to the dynamics of the
encoded BRN.
The obtained IG relates only components that have actual influence in the dynamics.
This typically results into IGs that contain less interaction than an hypothetical starting prior
IG, as the specification of component evolutions may reveal non-functional regulations.
Based on the derived IG, various static analysis allows to conclude on global properties
of the system dynamics (see \cite{PR11-SASB} for a short survey).
The most known are the Thomas' conjectures (that have been proved since \todo{refs}),
which relate the absence of
positive cycle to the impossibility for multi-stationarity (distinct attractors),
and the absence of negative cycle to the impossibility for oscillatory behaviours.
}
\todo{check if the inferred IG contains the IG of compatible Thomas}

\updated{%
Second, we tackle the characterisation of Thomas' parameters that are compatible with a given PH
model, \ie that fully specify the evolution of components while respecting the cooperations
specified in PH.
More formally, the resulting BRN dynamics is ensured to be included in the PH dynamics: any
transition in the (asynchronous) Thomas model exists in the PH model.
In practice, the number of compatible Thomas parameters can be huge for large networks where only a
few cooperations are specified.
This highlights the fact that the PH is very suited to analyse dynamics of large-scale networks
where the knowledge on cooperations is often incomplete:
instead of dealing with a possibly non-tractable number of Thomas models, one can already capture
precise dynamical features using a single PH models which gathers all the plausible dynamics.
}

\updated{%
Finally, we discuss on an implementation of these inference schemes using Answer Set
Programming (ASP) \cite{Baral03},  which turns out to be effective for these enumerative searches.
}

%The outcome of this work is twofold.
%The first benefit is that such an approach makes it possible to refine the construction of
%BRNs with a partial and progressively brought knowledge in PH, while being able to export such
%models in the Thomas' framework.
%This work thus strengthens the formal link between both modelings.
%The second feature of our method is that it can be applied on very large BRNs.

Our work is related to the approach of \cite{Khalis09} which relies on temporal logic, and \cite{20646302,DBLP:conf/ipcat/CorblinFTCT12} which also uses constraint programming.
Both aim at determining a class of models which are consistent with available partial data on the regulatory structure and dynamical properties.
%These classes are built in order to infer properties common to some studied models.
%In our approach, we intend to focus on the Thomas' parameters inference
Our method is based on a model rather than on constraints, which allows to define some properties on the system structure (such as cooperations).
Furthermore, we claim that we are able to deal with larger biological networks.
Finally, it must be noticed that we are not interested in this paper in the derivation of one PH
from a BRN (which was previously described in \cite{PMR10-TCSB}) but, on the contrary, to finding
out a set of BRNs from one PH.

The contributions presented in this paper significantly extends and improves the preliminary results
presented in \cite{FPIMR12-CMSB}.
In addition of improving the efficiency and accuracy of the IG inference, we have added support for
unsigned regulations, that are regulations being positive or negative depending on a particular
context.
This allows to apply our methodology to a wider class of models.
\todo{asp improvement? more detailed application?}

\paragraph{Outline}
\pref{sec:frameworks} recalls the PH and Thomas frameworks;
\pref{sec:infer-IG} defines the IG inference from PH;
\pref{sec:infer-K} details the enumeration of Thomas parametrizations compatible with a PH
and discuss its implementation in ASP.
\todo{add ref to ASP section}
\pref{sec:examples} illustrates the applicability of our method on simple examples
and large biological models.

### 2013-07-05 parts/ph.tex
avancées sur l'intro et l'abstract

\paragraph{Modeling cooperation}
\rewrite{Detail the modeling of cooperations?}
As described in \cite{PMR10-TCSB}, the cooperation between several processes to make another process bounce can be
expressed in PH by building a \emph{cooperative sort}.
\pref{fig:PH-cooperativity} shows an example of cooperation between processes $b_1$ and $c_1$ to
make $a_1$ bounce to $a_2$:
a cooperative sort $bc$ is defined with 4 processes (one for each sub-state of the presence of
processes $b_1$ and $c_1$).
For the sake of clarity, the processes of $bc$ are indexed using the sub-state they represent.
Hence, $bc_{01}$ represents the sub-state $\state{b_0,c_1}$, and so on.
Each process of sort $b$ and $c$ hit $bc$ to make it bounce to the process reflecting the status of the sorts $b$
and $c$ (e.g., $\PHfrappe{b_1}{bc_{00}}{bc_{10}}$ and $\PHfrappe{b_1}{bc_{01}}{bc_{11}}$).
Then, it is process $bc_{11}$ which hits $a_1$ to make it bounce to $a_2$ instead of
independent hits from $b_1$ and $c_1$.

@@

\rewrite{Explain the efficiency of PH for reachability?}

>>>>

\paragraph{Modeling cooperation}
%\rewrite{Detail the modeling of cooperations?}
As described in \cite{PMR10-TCSB}, the cooperation between several processes to make another process bounce can be
expressed in PH by building a \emph{cooperative sort}.
\pref{fig:PH-cooperativity} shows an example of cooperation between processes $b_1$ and $c_1$ to
make $a_1$ bounce to $a_2$:
a cooperative sort $bc$ is defined with 4 processes (one for each sub-state of the presence of
processes $b_1$ and $c_1$).
For the sake of clarity, the processes of $bc$ are indexed using the sub-state they represent.
Hence, $bc_{01}$ represents the sub-state $\state{b_0,c_1}$, and so on.
Each process of sort $b$ and $c$ hit $bc$ to make it bounce to the process reflecting the status of the sorts $b$
and $c$ (e.g., $\PHfrappe{b_1}{bc_{00}}{bc_{10}}$ and $\PHfrappe{b_1}{bc_{01}}{bc_{11}}$).
Then, it is process $bc_{11}$ which hits $a_1$ to make it bounce to $a_2$ instead of
independent hits from $b_1$ and $c_1$.

@@

%\rewrite{Explain the efficiency of PH for reachability?}
